Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2014-07-16 cordillera-cycle.tex
Change result and discussion headers.
Implement minor comments on section 4.1.

\todo{Do we need a new subsection title here?}

@@

% ----------------------------------------------------------------------
\section{Results}
\label{sec:results}
% ----------------------------------------------------------------------

Despite large differences in the input atmospheric forcing (temperature offset
time-series; Fig.~\ref{fig:timeseries}), model output presents consistent
features that can be observed across the range of forcing data used. In all
simulations, modelled ice volumes remain relatively low during most of the
glacial cycle, except during two major glacial events which occur between 61.9
and 55.4\,ka during Marine Oxygen Isotope Stage (MIS)~4, and between 29.5 and
16.9\,ka during MIS~2 (Fig.~\ref{fig:timeseries}). An ice volume minimum is
consistently attained between 52.2 and 45.6\,ka during MIS~3. However, the
magnitude and precise timing of these three events depend significantly on the
choice of proxy record used for time-dependent atmospheric forcing
(Table~\ref{tab:records})\renote{
    Hopefully this new table will assess several of your comments at once.}.

Simulations forced by the Greenland ice core palaeo-temperature\renote{
    Throughout this section, I prepended ``palaeo-temperature'' to
    ``record(s)'' in place of ``climate'' as you originally suggested. If we
    are to precise what type ``records'' we use, let's be clear about it.}
records (GRIP, NGRIP) produce the highest variability in modelled ice volume
throughout the last glacial cycle. In contrast, simulations driven by oceanic
(ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok) palaeo-temperature records
generally result in smaller\renote{
    Until we run out of other ajectives, I don't think that ``modest'' and
    ``generous'' are adequate terms for describing ice volumes.}
modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
More broadly speaking, they produce lower ice volume variability
throughout the simulation length. The NGRIP atmospheric forcing is the only one
that results in a larger ice volume during MIS~4 than during the last glacial
maximum (MIS~2).

@@

Despite such differences in the timing of attained volume maxima and mimima,
snapshots of model output show relatively consistent patterns of glaciation
(Fig.~\ref{fig:snapshots}). As a result of scaling factors applied to the
different palaeo-temperature proxy records (Table~\ref{tab:records}), modelled
ice sheet geometries during the last glacial maximum (MIS~2;
Fig.~\ref{fig:snapshots}, lower panels) appear highly similar from one
simulation to the next. They include a ca. 1\,500\,km-long central divide
located about 3\,500\,m\ above sea level (a.s.l.) along the spine of the Rocky
Mountains. The volumes of these modelled last glacial maximum ice sheets are
contained within a close range from 8.24 to 8.66\,m of sea-level equivalent
(s.l.e; Table~\ref{tab:records}).
A central ice cap persists in all simulations over the Skeena
Mountains during the MIS~3 ice volume minima (Fig.~\ref{fig:snapshots}, middle
panels). However, the size of the Skeena Mountains ice cap, and the dimensions
of the Cordilleran ice sheet during the MIS~4 glaciation, depend sensitively on
the choice of palaeo-temperature record used to drive the model.

% ----------------------------------------------------------------------
\section{Discussion}
\label{sec:discussion}
% ----------------------------------------------------------------------

\subsection{Timing of the glaciation peaks}

Independently of the choice of palaeo-temperature record used as forcing
time-series for the ice sheet model, our simulations consistently produce two
glaciation maxima during MIS~4 (61.9--55.4\,kyr) and MIS~2
(29.5--16.9\,kyr). These events broadly correspond in timing to the
Gladstone (MIS~4) and McConell (MIS~2) glaciations documented by geological
evidence in the northern part of the model domain \aref, and to the Fraser
(MIS~2) glaciation in the southern part \aref
\footnote{\textbf{Arjen, Martin, Johan,} can you check that I use these names
    correctly, and help me to fill the missing refs in this sentence? Is is
    possible to trace back when these names where first applied?}.

However, the timing of the modelled glacial maxima strongly depends on the
choice of palaeo-temperature record used, and a more precise agreement with
timing reconstructed from geological evidence is only attained in some of our
simulations. In the Puget lowland, the last glacial maximum advance of the
southern Cordilleran ice sheet margin has been constrained by radiocarbon
dating between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,kyr\,BP}
\citep[Fig.~4]{Porter.Swanson.1998}.
These dates have been confirmed by the offshore sediment record, which shows an
increase of glaciomarine sedimentation between 19.5 and
16.2\,\unit{\chem{^{14}C}\,cal\,kyr\,BP} \citep{Cosma.etal.2008}.

Among simulations presented here, only those forced by the GRIP, EPICA and
Vostok records place the last glacial maximum within this range
(Fig.~\ref{fig:timeseries}, lower panel). Simulations driven by the NGRIP,
ODP~1012 and ODP~1020 records, on the contrary, produce a last glacial maximum
that pre-dates reconstructed ages by several thousand years. Concerning the
simulations driven by oceanic records, this early deglaciation is caused by an
early warming sealed in the palaeo-temperature reconstructions
(Fig.~\ref{fig:timeseries}, upper panel). However, it has been shown that this
early warming is a local effect, corresponding to a weakening of the California
current \citep[Fig.~3]{Herbert.etal.2001}. The California current, driving cold
waters southwards
along the western coast of North America, is thought to have weakened in the
presence of the Cordilleran ice sheet, resulting in paradoxically warmer
sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
during the last glacial maximum \citep{Herbert.etal.2001}.

Because most of the Cordilleran ice sheet marine front lies in a different
oceanic circulation gyre than that holding the California current, it is
unlikely that it was affected by this early warming. However, the above paradox
illustrates the complexity of ice-sheet feedbacks on regional climate, and
demonstrates that, although located in the neighbourhood of the modelling
domain, the ODP~1012 and ODP~1020 palaeo-temperature proxy records can not be
used as a realistic forcing time-series to model the Cordilleran ice sheet
through the last glacial cycle.

These results make us believe that a palaeo-temperature proxy record located
in the direct vicinity of the former Cordilleran ice sheet would be of great
help to further understand its glacial history. Because, to our knowledge, no
such record yet exists, we focus the rest of our analysis on simulations forced
by temperature records from Greenland (GRIP) and Antarctic (EPICA) ice cores
(Fig.~\ref{fig:timeseries}, lower panel, dotted lines).

\subsection{Spatial imprint of ice cover}

@@

\subsection{Deglaciation history}

@@

% tab:records
\begin{table*}[t]
  \caption{Extrema of ice volume and extent corresponding to MIS~4, 3 and 2 for
           each of the low-resolution simulations (Fig.~\ref{fig:timeseries}).}
  \label{tab:records}
  {\begin{tabular}{l*{3}{|ccc}}
    \tophline
             & \multicolumn{3}{c}{Age (ka)}
             & \multicolumn{3}{c}{Ice extent (\unit{10^6\,km^2})}
             & \multicolumn{3}{c}{Ice volume (m~s.l.e.)} \\
    Record   & MIS.~4 & MIS.~3 & MIS.~2
             & MIS.~4 & MIS.~3 & MIS.~2
             & MIS.~4 & MIS.~3 & MIS.~2 \\
    \middlehline
    GRIP     & -57.58 & -49.24 & -19.52
             &   1.98 &   0.46 &   2.13
             &   7.52 &   0.89 &   8.52 \\
    NGRIP    & -60.26 & -50.16 & -22.85
             &   2.16 &   0.50 &   2.09
             &   8.69 &   0.93 &   8.24 \\
    EPICA    & -61.87 & -45.57 & -17.10
             &   1.57 &   0.95 &   2.08
             &   5.20 &   2.44 &   8.35 \\
    Vostok   & -60.87 & -49.68 & -16.87
             &   1.55 &   0.86 &   2.14
             &   5.10 &   2.01 &   8.66 \\
    ODP 1012 & -55.41 & -47.08 & -23.21
             &   1.44 &   0.85 &   2.13
             &   4.50 &   2.06 &   8.46 \\
    ODP 1020 & -60.16 & -52.24 & -29.46
             &   1.32 &   0.70 &   2.08
             &   3.88 &   1.52 &   8.32 \\
    \middlehline
    Minimum  & -61.87 & -52.24 & -29.46
             &   1.32 &   0.46 &   2.08
             &   3.88 &   0.89 &   8.24 \\
    Maximum  & -55.41 & -45.57 & -16.87
             &   2.16 &   0.95 &   2.14
             &   8.69 &   2.44 &   8.66 \\
    \bottomhline
  \end{tabular}}
  \belowtable{}
\end{table*}

% fig:locmap
\begin{figure}
  \includegraphics{locmap}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
           Innuitian (IIS) and Greenland (GIS) ice sheets during the last
           18\,\unit{\chem{^{14}C}\,kyr\,BP} (21.4\,cal\,kyr\,BP)
           \citep{Dyke.2004}. The rectangular box denotes the location of the
           modelling domain used in this study. Major mountain ranges covered
           by the ice sheet include the Alaska Range (AR), the Wrangell and
           St.-Elias mountains (WSE), the Selwyn and MacKenzie mountains (SMK),
           the Skeena Mountains (SM), the Coast Mountains (CM), the Rocky
           Mountains (RM) and the North Cascades (NC). The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2014}.
           \todo{Mark palaeo-ice sheets and mountain ranges on the map.}}
  \label{fig:locmap}
\end{figure}

>>>>

\noindent\todo{Do we need a new 2.1 subsection title here?}

@@

% ----------------------------------------------------------------------
\section{Sensitivity to atmospheric forcing time-series}
\label{sec:results}
% ----------------------------------------------------------------------

Despite large differences in the input atmospheric forcing (temperature offset
time-series; Fig.~\ref{fig:timeseries}), model output presents consistent
features that can be observed across the range of forcing data used. In all
simulations, modelled ice volumes remain relatively low during most of the
glacial cycle, except during two major glacial events which occur between 61.9
and 55.4\,ka during Marine Oxygen Isotope Stage (MIS)~4, and between 29.5 and
16.9\,ka during MIS~2 (Fig.~\ref{fig:timeseries}). An ice volume minimum is
consistently attained between 52.2 and 45.6\,ka during MIS~3. However, the
magnitude and precise timing of these three events depend significantly on the
choice of proxy record used for time-dependent atmospheric forcing
(Table~\ref{tab:extrema})\renote{
    Hopefully this new table will assess several of your comments at once.}.

Simulations forced by the Greenland ice core palaeo-temperature\renote{
    From here on, I prepend ``palaeo-temperature'' to ``record(s)'' in place of
    ``climate'' as you (Arjen) originally suggested. If we are to precise what
    type of ``records'' we use, let's be clear about it.}
records (GRIP, NGRIP) produce the highest variability in modelled ice volume
throughout the last glacial cycle. In contrast, simulations driven by oceanic
(ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok) palaeo-temperature records
generally result in smaller\renote{
    Until we run out of other ajectives, I don't think that ``modest'' and
    ``generous'' are adequate terms for describing ice volumes.}
modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
More broadly speaking, they produce lower ice volume variability
throughout the simulation length. The NGRIP atmospheric forcing is the only one
that results in a larger ice volume during MIS~4 than during the last glacial
maximum (MIS~2).

@@

Despite such differences in the timing of attained volume maxima and mimima,
snapshots of model output show relatively consistent patterns of glaciation
(Fig.~\ref{fig:snapshots}). As a result of scaling factors applied to the
different palaeo-temperature proxy records (Table~\ref{tab:records}), modelled
ice sheet geometries during the last glacial maximum (MIS~2;
Fig.~\ref{fig:snapshots}, lower panels) appear highly similar from one
simulation to the next. They include a ca. 1\,500\,km-long central divide
located about 3\,500\,m\ above sea level (a.s.l.) along the spine of the Rocky
Mountains. The volumes of these modelled last glacial maximum ice sheets are
contained within a close range from 8.24 to 8.66\,m of sea-level equivalent
(s.l.e; Table~\ref{tab:extrema}).
A central ice cap persists in all simulations over the Skeena
Mountains during the MIS~3 ice volume minima (Fig.~\ref{fig:snapshots}, middle
panels). However, the size of the Skeena Mountains ice cap, and the dimensions
of the Cordilleran ice sheet during the MIS~4 glaciation, depend sensitively on
the choice of palaeo-temperature record used to drive the model.

% ----------------------------------------------------------------------
\section[Comparison to the geologic record]
        {Comparison to the geologic record\renote{
    Arjen, you have proposed 2.1 Evaluation of model runs agains field
    evidence, 2.2 Palaeoglaciology of the Cordilleran ice sheet, and 2.3 The
    last deglaciation; but I don't think these headers work. In my vision of
    the paper, this entire ``discussion'' part is all about evaluation of model
    runs agains field evidence and palaeoglaciology. Because of this, and
    because we also present new ``results'', or rather different views of the
    same results, I propose to rename it to a more explicit ``Comparison to the
    geologic record'' (or similar). I have also renamed the ``results'' to
    ``Sensitivity to atmospheric forcing time-series'', which is actually what
    that section is about: raw model output, no discussion against the geologic
    record.}}
\label{sec:discussion}
% ----------------------------------------------------------------------

\subsection{Timing of glaciation}

Independently of the choice of palaeo-temperature record\renote{
    In my view, ``atmospheric/climate forcing'' would be far too broad here. We
    could possibly apply a much wilder choice of atmospheric forcing
    techniques, leading just as wild differences in model output
    \citep[e.g.,][but possibly much more than that]{Seguinot.etal.2014}.}
used to force the ice sheet model, our simulations consistently produce two
glacial maxima during the last glacial cycle. The first maximum configuration
is obtained during MIS~4 (61.9--55.4\,ka) and the second during MIS~2
(29.5--16.9\,ka). These events broadly correspond in timing to the Gladstone
(MIS~4) and McConnell (MIS~2) glaciations documented by geological evidence for
the northern sector of the Cordilleran ice sheet
    \citep{Duk-Rodkin.etal.1996, Ward.etal.2007,
           Stroeven.etal.2010, Stroeven.etal.2014},
and to the Fraser Glaciation (MIS~2) documented for its southern sector
    \citep{Porter.Swanson.1998, Margold.etal.inpress}.
There is patchy stratigraphical evidence for glaciations older than the Fraser
Glaciation \citep{Clague.Ward.2011} in British Columbia, and their extent and
timing are therefore still highly conjectural
    \citep[perhaps MIS~4 or early MIS~3; e.g.,][]{Cosma.etal.2008}.

However, the exact timing of modelled MIS~2 maxima depend strongly on the
choice of applied palaeo-temperature record, which allows for a more in-depth
comparison with geological evidence for the timing of maximum Cordilleran ice
sheet extent. In the Puget lowland (Fig.~\ref{fig:locmap}), the LGM advance of
the southern Cordilleran ice sheet margin has been constrained by radiocarbon
dating on wood between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,kyr\,BP}
\citep{Porter.Swanson.1998}.
These dates have been confirmed by the offshore sediment record, which shows an
increase of glaciomarine sedimentation between 19.5 and
16.2\,\unit{\chem{^{14}C}\,cal\,kyr\,BP} \citep{Cosma.etal.2008}. Radiocarbon
dating of the northern Cordilleran ice sheet margin is much less constrained
but straddles presented constraints from the southern margin. However,
cosmogenic exposure dating by \citep{Stroeven.etal.2010, Stroeven.etal.2014}
places the timing of maximum CIS extent during the McConnell glaciation close
to 17\,ka.

Among the simulations presented here, only those forced with the GRIP, EPICA
and Vostok palaeo-temperature records yield Cordilleran ice sheet maximum
extent that may be compatible with these field constraints
(Fig.~\ref{fig:timeseries}, lower panel; Fig.~\ref{fig:snapshots}, lowest
panels). Simulations driven by the NGRIP, ODP~1012 and ODP~1020
palaeo-temperature records, on the contrary, yield MIS~2 maximum Cordilleran
ice sheet extents that pre-date field-based constraints by several thousands of
years. Concerning the simulations driven by oceanic records, this early
deglaciation is caused by an early warming present in the alkenone
palaeo-temperature reconstructions (Fig.~\ref{fig:timeseries}, upper panel;
\citealp[Fig.3]{Herbert.etal.2001}). However, this
early warming is a local effect, corresponding to a weakening of the California
current \citep[Fig.~3]{Herbert.etal.2001}. The California current, driving cold
waters southwards along the south-western coast of North America\renote{
    Not only USA but also Baja California in Mexico.},
has been shown to have weakened during each peak global glaciation (in SPECMAP)
durint the past 550\,ka, resulting in paradoxically warmer sea-surface
temperatures at the locations of the ODP~1012 and ODP~1020 sites also during
the LGM \citep{Herbert.etal.2001}.

Because most of the marine margine of the Cordilleran ice sheet terminated in a
sector of the Pacific Ocean unaffected by variations in the California current,
it probably remained unaffected by this early warming. However, the above paradox
illustrates the complexity of ice-sheet feedbacks on regional climate, and
demonstrates that, although located in the neighbourhood of the modelling
domain, the ODP~1012 and ODP~1020 palaeo-temperature records cannot be
used as a realistic forcing to model the Cordilleran ice sheet
through the last glacial cycle.

\todo{say something about NGRIP.}

... Hence, we focus the rest of our analysis on simulations forced by palaeo-temperature records from GRIP and EPICA ice cores (Fig.~\ref{fig:timeseries}, lower panel, dotted lines).

\todo{avoid repetition with the next paragraph.}

These results indicate that a palaeo-temperature proxy record located in the
direct vicinity of the former Cordilleran ice sheet would potentially lead to
improved understanding of the history of advance and retreat of the western ice
sheet margin. Because, to our knowledge, no
such record yet exists, we focus the rest of our analysis on simulations forced
by palaeo-temperature records from GRIP and EPICA ice cores
(Fig.~\ref{fig:timeseries}, lower panel, dotted lines).

\subsection{Location and longevity of the major ice-despersal centres}

@@

\subsection{Pace and patterns of deglaciation}

@@

% tab:records
\begin{table*}[t]
  \caption{Extrema of ice volume and extent corresponding to MIS~4, 3 and 2 for
           each of the low-resolution simulations (Fig.~\ref{fig:timeseries}).}
  \label{tab:extrema}
  {\begin{tabular}{l*{3}{|ccc}}
    \tophline
             & \multicolumn{3}{c}{Age (ka)}
             & \multicolumn{3}{c}{Ice extent (\unit{10^6\,km^2})}
             & \multicolumn{3}{c}{Ice volume (m~s.l.e.)} \\
    Record   &  MIS~4 &  MIS~3 &  MIS~2
             &  MIS~4 &  MIS~3 &  MIS~2
             &  MIS~4 &  MIS~3 &  MIS~2 \\
    \middlehline
    GRIP     & -57.58 & -49.24 & -19.52
             &   1.98 &   0.46 &   2.13
             &   7.52 &   0.89 &   8.52 \\
    NGRIP    & -60.26 & -50.16 & -22.85
             &   2.16 &   0.50 &   2.09
             &   8.69 &   0.93 &   8.24 \\
    EPICA    & -61.87 & -45.57 & -17.10
             &   1.57 &   0.95 &   2.08
             &   5.20 &   2.44 &   8.35 \\
    Vostok   & -60.87 & -49.68 & -16.87
             &   1.55 &   0.86 &   2.14
             &   5.10 &   2.01 &   8.66 \\
    ODP 1012 & -55.41 & -47.08 & -23.21
             &   1.44 &   0.85 &   2.13
             &   4.50 &   2.06 &   8.46 \\
    ODP 1020 & -60.16 & -52.24 & -29.46
             &   1.32 &   0.70 &   2.08
             &   3.88 &   1.52 &   8.32 \\
    \middlehline
    Minimum  & -61.87 & -52.24 & -29.46
             &   1.32 &   0.46 &   2.08
             &   3.88 &   0.89 &   8.24 \\
    Maximum  & -55.41 & -45.57 & -16.87
             &   2.16 &   0.95 &   2.14
             &   8.69 &   2.44 &   8.66 \\
    \bottomhline
  \end{tabular}}
  \belowtable{}
\end{table*}

% fig:locmap
\begin{figure}
  \includegraphics{locmap}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
           Innuitian (IIS) and Greenland (GIS) ice sheets during the last
           18\,\unit{\chem{^{14}C}\,kyr\,BP} (21.4\,cal\,kyr\,BP)
           \citep{Dyke.2004}. The rectangular box denotes the location of the
           modelling domain used in this study. Major mountain ranges covered
           by the ice sheet include the Alaska Range (AR), the Wrangell and
           St.-Elias mountains (WSE), the Selwyn and MacKenzie mountains (SMK),
           the Skeena Mountains (SM), the Coast Mountains (CM), the Rocky
           Mountains (RM) and the North Cascades (NC). The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2014}.
           \todo{Mark palaeo-ice sheets and mountain ranges on the map.
                 Mark location of the Puget Lowland.}}
  \label{fig:locmap}
\end{figure}
